---
title: "Stat405 Bayesian Coding Part"
author: "Aiwei Yang42274209, Chang Sun 19062439"
date: "2026-04-17"
output: pdf_document
---

```{r setup, include=FALSE}
# Load required packages
library(readr)
library(dplyr)
library(ggplot2)
library(cmdstanr)
library(posterior)
library(bayesplot)

theme_set(theme_minimal(base_size = 12))
set.seed(123)
```

1.  Read and prepare data

```{r}
# Read dataset
df <- read_csv("day2018.csv", show_col_types = FALSE)

# Convert selected variables to factors
bike_clean <- df %>%
  mutate(
    season = factor(season),
    yr = factor(yr),
    workingday = factor(workingday),
    holiday = factor(holiday)
  )
# Standardize continuous predictors
bike_clean <- bike_clean %>%
  mutate(
    temp_std = as.numeric(scale(temp)),
    hum_std = as.numeric(scale(hum)),
    windspeed_std = as.numeric(scale(windspeed))
  )
# Quick check
glimpse(bike_clean)
summary(bike_clean)
```

2.  Build design matrix

```{r}
X <- model.matrix(
  cnt ~ season + yr + temp_std + hum_std + windspeed_std + workingday + holiday,
  data = bike_clean
)
dim(X)
head(X)
```

3.  Prepare data for Stan

```{r}
stan_data <- list(
  N = nrow(X),
  K = ncol(X),
  y = bike_clean$cnt,
  X = X
)

stan_data$N
stan_data$K
head(stan_data$y)
```

4.  Remove old stan file if they exist

```{r}
if (file.exists("poisson_model.stan")) file.remove("poisson_model.stan")
if (file.exists("negbin_model.stan")) file.remove("negbin_model.stan")
```

5.  Write Poisson Stan model

```{r}
poisson_code <- "
data {
  int<lower=1> N;
  array[N] int<lower=0> y;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  vector[K] beta;
}
model {
  beta ~ normal(0, 2.5);
  y ~ poisson_log(X * beta);
}
generated quantities {
  vector[N] log_lik;
  array[N] int y_rep;

  for (n in 1:N) {
    log_lik[n] = poisson_log_lpmf(y[n] | X[n] * beta);
    y_rep[n] = poisson_log_rng(X[n] * beta);
  }
}
"

writeLines(poisson_code, "poisson_model.stan")

# Verify the file contents
readLines("poisson_model.stan")
```

6.  Compile Poisson model

```{r}
poisson_mod <- cmdstan_model("poisson_model.stan")
```

7.  Fit Poisson model using HMC

```{r}
fit_poisson_hmc <- poisson_mod$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1500,
  iter_sampling = 1500,
  adapt_delta = 0.99,
  max_treedepth = 15,
  seed = 123,
  refresh = 500
)  

poisson_hmc_summary <- fit_poisson_hmc$summary()
poisson_hmc_summary
```

8.  Inspect Poisson HMC diagnostics

```{r}
# Print summary table
poisson_hmc_summary

# Check the largest R-hat value
max(poisson_hmc_summary$rhat, na.rm = TRUE)

# Check the smallest bulk ESS
min(poisson_hmc_summary$ess_bulk, na.rm = TRUE)

# Check the smallest tail ESS
min(poisson_hmc_summary$ess_tail, na.rm = TRUE)

# Traceplots for selected Poisson coefficients
mcmc_trace(
  fit_poisson_hmc$draws(variables = c("beta[1]", "beta[2]", "beta[3]"))
)
```

9.  Fit Poisson model using Variational Inference

```{r}
fit_poisson_vi <- poisson_mod$variational(
  data = stan_data,
  algorithm = "meanfield",
  iter = 20000,
  output_samples = 2000,
  seed = 123
)

# Extract summary
poisson_vi_summary <- fit_poisson_vi$summary()

# Print summary table
poisson_vi_summary
```

10. Compare Poisson HMC and VI

```{r}
poisson_hmc_beta <- fit_poisson_hmc$summary(variables = "beta")
poisson_vi_beta  <- fit_poisson_vi$summary(variables = "beta")

comparison_poisson <- data.frame(
  parameter = poisson_hmc_beta$variable,
  hmc_mean  = poisson_hmc_beta$mean,
  hmc_sd    = poisson_hmc_beta$sd,
  vi_mean   = poisson_vi_beta$mean,
  vi_sd     = poisson_vi_beta$sd
)

comparison_poisson
```

11. Write Negative Binomial Stan model

```{r}
negbin_code <- "
data {
  int<lower=1> N;
  array[N] int<lower=0> y;
  int<lower=1> K;
  matrix[N, K] X;
}
parameters {
  vector[K] beta;
  real log_phi;
}
transformed parameters {
  real<lower=0> phi;
  phi = exp(log_phi);
}
model {
  beta ~ normal(0, 2.5);
  log_phi ~ normal(0, 1);
  y ~ neg_binomial_2_log(X * beta, phi);
}
generated quantities {
  vector[N] log_lik;
  array[N] int y_rep;
  for (n in 1:N) {
    log_lik[n] = neg_binomial_2_log_lpmf(y[n] | X[n] * beta, phi);
    y_rep[n] = neg_binomial_2_log_rng(X[n] * beta, phi);
  }
}
"

# Remove old file if it exists
if (file.exists("negbin_model.stan")) file.remove("negbin_model.stan")

# Write the Stan file
writeLines(negbin_code, "negbin_model.stan")

# Verify file contents
readLines("negbin_model.stan")
```

12. Compile Negative Binomial model

```{r}
nb_mod <- cmdstan_model("negbin_model.stan", force_recompile = TRUE)
```

13. Fit Negative Binomial model using HMC

```{r}
fit_nb_hmc <- nb_mod$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1500,
  iter_sampling = 1500,
  adapt_delta = 0.99,
  max_treedepth = 15,
  seed = 123,
  refresh = 500
)
# Extract summary
nb_hmc_summary <- fit_nb_hmc$summary()

# Print summary table
nb_hmc_summary
```

Since there is warning, we need to check r-hat and ESS

```{r}
nb_hmc_summary <- fit_nb_hmc$summary()

max(nb_hmc_summary$rhat)
min(nb_hmc_summary$ess_bulk)
```

14. Inspect Negative Binomial HMC diagnostics

```{r}
# Check the largest R-hat value
max(nb_hmc_summary$rhat, na.rm = TRUE)

# Check the smallest bulk ESS
min(nb_hmc_summary$ess_bulk, na.rm = TRUE)

# Check the smallest tail ESS
min(nb_hmc_summary$ess_tail, na.rm = TRUE)

# Traceplots for selected coefficients and dispersion parameter
mcmc_trace(
  fit_nb_hmc$draws(variables = c("beta[1]", "beta[2]", "beta[3]", "phi"))
)
```

15. Fit Negative Binomial model using Variational Inference

```{r}
fit_nb_vi <- nb_mod$variational(
  data = stan_data,
  algorithm = "meanfield",
  iter = 20000,
  output_samples = 2000,
  seed = 123
)

# Extract summary
nb_vi_summary <- fit_nb_vi$summary()

# Print summary table
nb_vi_summary
```

16. Compare Negative Binomial HMC and VI

```{r}
nb_hmc_beta <- fit_nb_hmc$summary(variables = c("beta", "phi"))
nb_vi_beta  <- fit_nb_vi$summary(variables = c("beta", "phi"))

comparison_nb <- data.frame(
  parameter = nb_hmc_beta$variable,
  hmc_mean  = nb_hmc_beta$mean,
  hmc_sd    = nb_hmc_beta$sd,
  vi_mean   = nb_vi_beta$mean,
  vi_sd     = nb_vi_beta$sd
)

comparison_nb
```
17. Posterior predictive checks
```{r}
# Extract posterior predictive draws
yrep_poisson <- as_draws_matrix(fit_poisson_hmc$draws("y_rep"))
yrep_nb <- as_draws_matrix(fit_nb_hmc$draws("y_rep"))

# Posterior predictive density overlay
ppc_dens_overlay(y = bike_clean$cnt, yrep = yrep_poisson[1:100, ])
ppc_dens_overlay(y = bike_clean$cnt, yrep = yrep_nb[1:100, ])

# Compare observed and replicated mean and variance
ppc_stats <- data.frame(
  model = c("Observed", "Poisson_HMC", "NB_HMC"),
  mean = c(
    mean(bike_clean$cnt),
    mean(rowMeans(yrep_poisson)),
    mean(rowMeans(yrep_nb))
  ),
  variance = c(
    var(bike_clean$cnt),
    mean(apply(yrep_poisson, 1, var)),
    mean(apply(yrep_nb, 1, var))
  )
)

ppc_stats
```

18. Save all summary tables

```{r}
write.csv(as.data.frame(poisson_hmc_summary), "poisson_hmc_summary.csv", row.names = FALSE)
write.csv(as.data.frame(poisson_vi_summary), "poisson_vi_summary.csv", row.names = FALSE)
write.csv(as.data.frame(nb_hmc_summary), "nb_hmc_summary.csv", row.names = FALSE)
write.csv(as.data.frame(nb_vi_summary), "nb_vi_summary.csv", row.names = FALSE)

write.csv(comparison_poisson, "comparison_poisson.csv", row.names = FALSE)
write.csv(comparison_nb, "comparison_nb.csv", row.names = FALSE)
write.csv(ppc_stats, "ppc_stats.csv", row.names = FALSE)
```

19. Compare computation times

```{r}
fit_poisson_hmc$time()
fit_poisson_vi$time()
fit_nb_hmc$time()
fit_nb_vi$time()
```
